---
title: "Statistical Computing Quiz 2022"
author: "Zeynep Filiz EREN"
date: "4/23/2022"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--IMPORTANT!!!!! Please fill out the sections that say:"write your ... here"-->

<br>

## **Questions**

<br>

### **<span style="text-decoration:underline">Question 1.</span>	**
**One way to measure a person’s fitness is to measure their body fat percentage. Average body fat percentages vary by age, but according to some guidelines, the normal range for men is 15-20% body fat, and the normal range for women is 20-25% body fat. <br> Our sample data is from a group of men and women who did workouts at a gym three times a week for a year. Then, their trainer measured the body fat and wonders whether the body fat is the same for men and women. The table below shows the data. **

<br>

|       |   |   |    |    |    |    |    |   |    |    |
|-------|---|---|----|----|----|----|----|---|----|----|
| Men   | 1 | 8 | 14 | 13 | 20 | 24 | 16 | 7 | 6  | 19 |
| Women | 5 | 7 | 22 | 21 | 23 | 28 | 16 | 6 | 12 | 23 |
|  |  |  |  |  |  |  |  |  |  |  |

<br>

<br>

```{r}
  men <- c(1,8,14,13,20,24,16,7,6,19)
  women <- c(5,7,22,21,23,28,16,6,12,23)
```

<br>

<span style="text-decoration:underline"> **a.**	Build histograms and boxplots for each group. Comment about the distribution and the outliers.</span>

<br>

```{r}
  hist(men,col="blue",breaks = 8)
  hist(women,col = "red",breaks = 8)
  
  boxplot(men)
  boxplot(women)
```

<br>

Men dataset is distributed normally but women dataset is not distributed normally.

<br>

<span style="text-decoration:underline"> **b.**	Write down a hypothesis to test if gender affects the amount of body fat. </span>

<br>

µ0 = mean(men)

µ1 = mean(women)

H0: µ0 = µ1

H1: µ0 != µ1

<br>

<span style="text-decoration:underline"> **c.**	Decide whether this is an independent or dependent two sample test.</span>

<br>

These datasets are not dependent, therefore we will do independent two sample t-test.

<br>

<span style="text-decoration:underline"> **d.**	Check normality assumption/ assumptions needed.</span>

<br>

```{r}
  shapiro.test(men)
  shapiro.test(women)
```

<br>

H0: Data is Normally Distributed

H1: Data is not Normally Distributed    alpha=0.05

If p <= 0.05: then the null hypothesis can be rejected (meaning the variable is not normally distributed).

If p > 0.05: then the null hypothesis can not be rejected (meaning the variable might be normally distributed or not).

p>0.05, two samples are normally distributed.

<br>

<span style="text-decoration:underline"> **e.**	Test your hypothesis. (α=0.05)</span>

<br>

```{r}
  t.test(men,women,alternative = "two.sided",mu=0,paired = FALSE,var.equal = TRUE)
```

<br>

<span style="text-decoration:underline"> **f.**	Comment:</span>

<br>

We can not reject H0, which means the means of the two datasets can be equal but they might be different as well.

<br>

<span style="text-decoration:underline"> **g.**	Write down a hypothesis to test if being a women means “the person is more likely to have more body fat”.</span>

<br>

µ0 = men

µ1 = women

H0: µ0 >= µ1

H1: µ0 < µ1

<br>

<span style="text-decoration:underline"> **h.**	Test your hypothesis. (α=0.05)</span>

<br>

```{r}
  t.test(men,women,alternative = "greater",mu=0,paired = FALSE,var.equal = TRUE)
```

<br>

<span style="text-decoration:underline"> **i.**	Comment:</span>

<br>

We can reject H0, which means women probably have more body fat compared to men.

<br>

<span style="text-decoration:underline"> **j.**	Find the confidence intervals of mean body fat for men and women separately and compare them. (α=0.05)</span>

<br>

```{r}
  t.score <-  qt(p=0.05, df=length(men)-1,lower.tail=F)

  men.mean <- mean(men)
  men.sd <- sd(men)
  men.se <- men.sd/sqrt(length(men))
  error <- t.score * men.se
  lower.bound <- men.mean  - error
  upper.bound <- men.mean  + error
  print(c(lower.bound,upper.bound))


  t.score <-  qt(p=0.05, df=length(women)-1,lower.tail=F)

  women.mean <- mean(women)
  women.sd <- sd(women)
  women.se <- women.sd/sqrt(length(women))
  error <- t.score * women.se
  lower.bound <- women.mean  - error
  upper.bound <- women.mean  + error
  print(c(lower.bound,upper.bound))
```

<br>



<br>

### **<span style="text-decoration:underline">Question 2.</span>	**
**The same data, but totally different scenario: Lets say the below table corresponds to the “%body fat of 10 people”, which are measured at the time they started doing sports for 3 times a week; and after 3 months. **

<br>

|       |   |   |    |    |    |    |    |   |    |    |
|-------|---|---|----|----|----|----|----|---|----|----|
| After 3 months | 1 | 8 | 14 | 13 | 20 | 24 | 16 | 7 | 6  | 19 |
| Start day      | 5 | 7 | 22 | 21 | 23 | 28 | 16 | 6 | 12 | 23 |
|  |  |  |  |  |  |  |  |  |  |  |

<br>

<br>

```{r}
  start <- c(5,7,22,21,23,28,16,6,12,23)
  after <- c(1,8,14,13,20,24,16,7,6,19)
```

<br>

<span style="text-decoration:underline">**a.**	Write down a hypothesis to test if doing sports decreases the amount of body fat.</span>

<br>

µ0 = start

µ1 = after

H0: µ0 <= µ1

H1: µ0 > µ1

<br>

<span style="text-decoration:underline">**b.**	Decide whether this is an independent or dependent two sample test.</span>

<br>

Since people are the same, samples are dependent. Therefore we will need dependent two sample t-test.

<br>

<span style="text-decoration:underline">**c.**	Check normality assumption/assumptions.</span>

<br>

```{r}
  shapiro.test(start)
  shapiro.test(after)
```

<br>

H0: Data is Normally Distributed
H1: Data is not Normally Distributed    alpha=0.05

<br>

<span style="text-decoration:underline">**d.**	Test your hypothesis. (α=0.05)</span>

<br>

```{r}
  t.test(start,after, mu =0 ,paired = TRUE, alternative = "less",conf.level = 0.95)
```

<br>

<span style="text-decoration:underline">**e.**	Comment:</span>

<br>

  We can not reject H0 and we can say that doing sports decreases the amount of body fat. 

<br>



<br>

### **<span style="text-decoration:underline">Question 3.</span>	**
**Another scenario: Lets say the below table corresponds to the “muscle gain of 30 people who started doing sports 6 months ago”: 1st group did sports once a week, 2nd group did sports three times a week, 3rd group did sports five times a week.  **



<br>

|       |   |   |    |    |    |    |    |   |    |    |
|-------|---|---|----|----|----|----|----|---|----|----|
| Group 1        | 1        | 3        | 6        | 5        | 7        | 2        | 1        | 2        | 3        | 5        |
|     Group 2    |     9    |     7    |     7    |     4    |     3    |     6    |     5    |     4    |     6    |     7    |
|     Group 3    |     6    |     8    |     7    |     7    |     9    |     5    |     7    |     8    |     6    |     7    |
|  |  |  |  |  |  |  |  |  |  |  |

<br>

<br>

```{r}
  group1 <- c(1,3,6,5,7,2,1,2,3,5)
  group2 <- c(9,7,7,4,3,6,5,4,6,7)
  group3 <- c(6,8,7,7,9,5,7,8,6,7)
```

<br>

<span style="text-decoration:underline">**a.**	Write down a hypothesis to test if these groups have equal means or not. (α=0.05)</span>

<br>

H0: µ0 = µ1 = µ2

H1: At least one of the means is different.

<br>

```{r}
groups <- c(group1,group2,group3)
frequent  <- as.factor(c(rep("once",10),rep("three",10),rep("five",10)))
data <- data.frame(groups,frequent)

bartlett.test(groups~frequent,data = data)

model <- aov(groups~frequent,data = data)
summary(model)
```

<br>

<span style="text-decoration:underline">**b.**	Comment:</span>

<br>

Since p>0.05, we can't reject H0. Therefore, we don't have any idea if the means are same.

<br>

<span style="text-decoration:underline">**c.**	Implement TukeyHSD test and compare the groups. (α=0.05)</span>

<br>

```{r}
  TukeyHSD(model)
```

<br>

<span style="text-decoration:underline">**d.**	Comment: </span>

<br>

Since p-values of the comparisons "once-five" and "three-once" are less then 0.05, we can say that "once" is especially different from other two. 

<br>

<br>

<br>